---
engine: knitr
---

# GWAS with PLINK2

This tutorial demonstrates how to perform a **genome-wide association study (GWAS)** using **PLINK2**, with and without **principal component analysis (PCA)** adjustment.  
It also shows how to visualize results using the **`qqman`** and **`ggplot2`** R packages.

```{bash, echo=FALSE}
[ -f data/gwas.imputed.chr20.dose.vcf.gz ] || wget -q -P data https://genepi.i-med.ac.at/downloads/imputation/gwas.imputed.chr20.dose.vcf.gz
mkdir -p output
```

## GWAS Without PCA

First, we run a logistic regression GWAS on a simulated phenotype without adjusting for population structure.

#### Run GWAS

```{bash, results='hide'}
plink2 --vcf data/gwas.imputed.chr20.dose.vcf.gz \
  --pheno data/phenotypes.txt \
  --pheno-name pheno_1 \
  --logistic allow-no-covars \
  --out output/gwas
```

**Explanation:**

* `--vcf`: Input imputed genotype file in VCF format
* `--pheno`: Phenotype file (tab-delimited)
* `--pheno-name`: Selects the phenotype column to test (here `pheno_1`)
* `--logistic allow-no-covars`: Runs logistic regression without covariates
* `--out`: Prefix for output files


#### Visualize GWAS Results

```{r}
library(qqman)

pheno_1 <- read.table("output/gwas.pheno_1.glm.logistic.hybrid", header = TRUE, comment.char = "")
pheno_1 <- pheno_1[!is.na(pheno_1$P), ]

# Manhattan plot
manhattan(x = pheno_1, chr = "X.CHROM", bp = "POS", p = "P", snp = "ID")

# QQ plot
qq(pheno_1$P)
```

**Notes:**

* The `qqman` package provides quick visualizations of GWAS results.
* The Manhattan plot highlights SNPs across chromosome 20, while the QQ plot checks for p-value inflation.

## PCA: Assessing Population Structure

Before including principal components (PCs) as covariates, let's visualize the PCA data.

```{r}
library(ggplot2)

pca_data <- read.table("data/covariates.txt", header = TRUE, comment.char = "")

ggplot(pca_data, aes(x = PC1, y = PC2)) +
  geom_point(alpha = 0.7) +
  theme_minimal() +
  labs(title = "PCA of Genotype Data", x = "PC1", y = "PC2")
```

**Explanation:**

* `covariates.txt` contains individual-level covariates including PCs (e.g., `IID`, `PC1`, `PC2`, …).
* The PCA plot can reveal population substructure or outliers.

## GWAS With PCA Adjustment

Next, include the PCA covariates in the GWAS to account for population stratification.

#### Run GWAS with Covariates

```{bash, results='hide'}
plink2 --vcf data/gwas.imputed.chr20.dose.vcf.gz \
  --pheno data/phenotypes.txt \
  --pheno-name pheno_1 \
  --covar data/covariates.txt \
  --logistic hide-covar \
  --out output/gwas.covar
```

**Explanation:**

* `--covar`: Provides a file containing covariates (e.g., PCs).
* `--logistic hide-covar`: Runs logistic regression while hiding covariate results in the output.

#### Visualize Adjusted GWAS Results

```{r}
pheno_1 <- read.table("output/gwas.covar.pheno_1.glm.logistic.hybrid", header = TRUE, comment.char = "")
pheno_1 <- pheno_1[!is.na(pheno_1$P), ]

# Manhattan plot
manhattan(x = pheno_1, chr = "X.CHROM", bp = "POS", p = "P", snp = "ID")

# QQ plot
qq(pheno_1$P)
```

**Comparison Tip:**
Compare the QQ plots *before* and *after* including PCs to see how population structure adjustment affects the inflation of p-values.

---

## References

* [PLINK2 Documentation](https://www.cog-genomics.org/plink/2.0/)
* [qqman R Package](https://cran.r-project.org/package=qqman)
* [ggplot2 R Package](https://ggplot2.tidyverse.org/)
